---
title: "The full pipeline"
subtitle: "Load, preprocess, tune, test, and predict, once, end to end, before any of it gets explained."
author: Franck Albinet
date: last-modified
format:
  html:
    toc: false
---

This page runs the whole pipeline once, start to finish: load spectra, preprocess them, tune and fit a model, test it, predict with it. Every step below gets its own full page later in the course. This page exists to show the shape of the whole thing before any of that detail.

The ring trial (RT) samples used throughout this course come from a real interlaboratory study: labs in several countries, including the IAEA's, scanned the same soil samples on their own instruments, to see how much spectra disagree from lab to lab. @safanelli2023 covers what came out of it.

```{r}
# --- Config ------------------------------------------------------------------
library(here)
library(tidyverse)
library(tidymodels)
library(plsmod)
source(here("R", "prepare_data.R"))
source(here("R", "steps.R"))
source(here("R", "split.R"))
tidymodels_prefer()

SPECTRA_DIR = here("data/RT_spectra")
SOIL_PROP_FILE = here("data/RT_wetchem_soildata.csv")
TARGET <- "carbon_tot_perc"
RESAMPLING_GRID <- seq(3995, 600, by = -5)
```


```{r}
# --- Loading ------------------------------------------------------------------
# RT_44 dropped as a known outlier -> 5-outlier-detection.qmd
data <- prepare_rt(SPECTRA_DIR, SOIL_PROP_FILE) |> filter(smp_id != "RT_44")
```

```{r}
# --- Preprocessing --------------------------------------------------------------
# log-scale, resample, SNV, Savitzky-Golay -> 4-feature-engineering.qmd
spc_cols <- names(data)[startsWith(names(data), "wn_")]
meta_cols <- setdiff(names(data), c(TARGET, spc_cols))

rec <- recipe(data) |>
  update_role(all_of(TARGET), new_role = "outcome") |>
  update_role(all_of(spc_cols), new_role = "predictor") |>
  update_role(all_of(meta_cols), new_role = "metadata") |>
  step_log(all_outcomes(), base = 10) |>
  step_resample(grid = RESAMPLING_GRID) |>
  step_snv() |>
  step_sg(w = 15L, p = 3L, m = 1L)

# --- Tuning -----------------------------------------------------------------------
# search num_comp, pick the best, lock it in -> 6-evaluating-performance.qmd
split <- kennard_stone_split(data, prop = 50/70, cols = starts_with("wn_"))
folds <- vfold_cv(training(split), v = 10, repeats = 10, strata = TARGET, breaks = 2)

wf <- workflow() |>
  add_recipe(rec) |>
  add_model(pls(num_comp = tune()) |> set_engine("mixOmics") |> set_mode("regression"))

tune_res <- tune_grid(wf, resamples = folds, grid = tibble(num_comp = 2:20))
best <- select_best(tune_res, metric = "rmse")
final_wf <- wf |> finalize_workflow(best)

# --- Testing ------------------------------------------------------------------------
# fit on train, score once on the held-out test split
final_res <- final_wf |> last_fit(split)
collect_metrics(final_res) |> select(.metric, .estimate)

# --- Predicting ---------------------------------------------------------------------
# .pred and TARGET are both log10 here -> 2-loading-data.qmd, 3-eda.qmd
collect_predictions(final_res) |> 
  select(all_of(TARGET), .pred) |>
  head()
```

