---
title: "The full pipeline"
subtitle: "Load, preprocess, tune, test, and predict, once, end to end, before any of it gets explained."
author: Franck Albinet
date: last-modified
format:
  html:
    toc: true
---

This page runs the whole pipeline once, start to finish: load spectra, preprocess them, tune and fit a model, test it, predict with it. Every step below gets its own full page later in the course. This page exists to show the shape of the whole thing before any of that detail.

The RT (ring trial) samples used throughout this course come from a real interlaboratory study: labs in several countries, including the IAEA's, scanned the same soil samples on their own instruments, to see how much spectra disagree from lab to lab. The [paper covering it](https://www.sciencedirect.com/science/article/pii/S0016706123004019) was published in *Geoderma* in 2023. Some of its coauthors are colleagues on this course.

```{r}
#| echo: false
library(here)
library(tidyverse)
library(tidymodels)
library(plsmod)
source(here("R", "prepare_data.R"))
source(here("R", "steps.R"))
source(here("R", "split.R"))
theme_set(theme_bw())
tidymodels_prefer()
```

```{r}
SPECTRA_DIR = here("data/RT_spectra")
SOIL_PROP_FILE = here("data/RT_wetchem_soildata.csv")
TARGET <- "carbon_tot_perc"
RESAMPLING_GRID <- seq(3995, 600, by = -5)
```
## Loading
`prepare_rt()` reads the OPUS files and the wet chemistry into one tibble. One sample, `RT_44`, is dropped as a known outlier; `5-outlier-detection.qmd` covers how that gets found.

```{r}
data <- prepare_rt(SPECTRA_DIR, SOIL_PROP_FILE) |> filter(smp_id != "RT_44")
```

## Preprocessing

- log-scale the outcome, 
- resample onto a common wavenumber grid, 
- correct scatter with SNV, smooth and,
- smooth and differentiate with Savitzky-Golay. 

`4-feature-engineering.qmd` explains each one.

```{r}
spc_cols <- names(data)[startsWith(names(data), "wn_")]
meta_cols <- setdiff(names(data), c(TARGET, spc_cols))

rec <- recipe(data) |>
  update_role(all_of(TARGET), new_role = "outcome") |>
  update_role(all_of(spc_cols), new_role = "predictor") |>
  update_role(all_of(meta_cols), new_role = "metadata") |>
  step_log(all_outcomes(), base = 10) |>
  step_resample(grid = RESAMPLING_GRID) |>
  step_snv() |>
  step_sg(w = 15L, p = 3L, m = 1L)
```

## Tuning 

A PLS model's main hyperparameter is how many components it keeps.

- `tune()` marks that as unknown
- `tune_grid()` searches 19 candidate values across 100 resamples
- `select_best()` picks the one with the best average score
- `finalize_workflow()` locks it in.

```{r}
split <- kennard_stone_split(data, prop = 50/70, cols = starts_with("wn_"))
folds <- vfold_cv(training(split), v = 10, repeats = 10, strata = TARGET, breaks = 2)

wf <- workflow() |>
  add_recipe(rec) |>
  add_model(pls(num_comp = tune()) |> set_engine("mixOmics") |> set_mode("regression"))

tune_res <- tune_grid(wf, resamples = folds, grid = tibble(num_comp = 2:20))
best <- select_best(tune_res, metric = "rmse")
final_wf <- wf |> finalize_workflow(best)
best
```

## Testing

`last_fit()` fits the finalized workflow on the training split and scores it once against the test split it has never seen.

```{r}
final_res <- final_wf |> last_fit(split)
collect_metrics(final_res)
```

## Predicting
Every one of those predictions already happened inside `last_fit()`, on data the model never saw while fitting. `.pred` and `carbon_tot_perc` are both log10 values here, since the model was trained on the log-scaled outcome; `2-loading-data.qmd` and `3-eda.qmd` cover why that matters.

```{r}
collect_predictions(final_res) |> select(all_of(TARGET), .pred)
```

